---
title: "Liana and tree tissues chemistry"
author: "[**_Gbadamassi G.O. Dossa_**](https://scholar.google.com/citations?user=w20AkZcAAAAJ&hl=en)"
date: (Created on Jan 3 2022 and last updated on `r format(Sys.time(), "%a %b %d %Y %H:%M:%S")`)
output: 
#rmdformats::material
  html_document:
    fig_caption: yes #Enabling figure caption
    theme: spacelab #readable #sandstone #spacelab #flatly
    highlight: pygments #tango #kate
    toc: TRUE # Enabling table of content (toc) on the left side of output
    toc_depth: 3 # Showing toc up to 3 headings below (i.e.,1, 1.1, 1.1.1)
    number_sections: TRUE # Number sections
    toc_float:
      collapsed: TRUE # Allows the toc to collapse if the reading section is not concerned
      smooth_scroll: TRUE # Allows smoth scrolling of toc
      toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

Here, we are reporting the differences in chemical content of different tissues/organs from lianas and not lianas (self-supporting trees).  
This chemistry here is the initial quality of both wood and bark of five species of lianas and five species of trees with a dbh of 2.5 cm on average. We used logs from those ten species to set up an experiment in the secondary rainforest of XTBG in December 2021. 

# Input chemistry data

## Setting up data

```{r, setup_data}
library(tidyverse)
ini_chem<-readr::read_csv("liana wood bark ten species initial chemistry.csv")
head(ini_chem)
summary(ini_chem)
ini_chem$ID<-as.factor(ini_chem$ID)
ini_chem$Species<-as.factor(ini_chem$Species)
ini_chem$growth_form<-as.factor(ini_chem$growth_form)
ini_chem$Tissue<-as.factor(ini_chem$Tissue)
ini_chem$Position<-as.factor(ini_chem$Position)
```

## Compute chemical traits average per tissue

```{r}
ini_chem_tissue<-ini_chem%>%
  group_by(Species, growth_form, Tissue)%>%
  dplyr::summarise(mean_C=mean(C_percent, na.rm=T),
                   mean_N=mean(N_percent, na.rm=T),
                   mean_P=mean(P_percent, na.rm=T),
                   mean_Mg=mean(Mg_percent, na.rm=T),
                   mean_Ca=mean(Ca_percent, na.rm=T),
                   mean_Lignin=mean(Lignin_percent, na.rm=T),
                   mean_Tannin=mean(Condensed_tannin_percent, na.rm=T)
)
ini_chem_tissue
```


```{r}
ini_chem_tissue2<-ini_chem%>%
  group_by(growth_form, Tissue)%>%
  dplyr::summarise(mean_C=mean(C_percent, na.rm=T),
                   mean_N=mean(N_percent, na.rm=T),
                   mean_P=mean(P_percent, na.rm=T),
                   mean_Mg=mean(Mg_percent, na.rm=T),
                   mean_Ca=mean(Ca_percent, na.rm=T),
                   mean_Lignin=mean(Lignin_percent, na.rm=T),
                   mean_Tannin=mean(Condensed_tannin_percent, na.rm=T)
)
ini_chem_tissue2
```

# Update calculation and calculations of means and SD with new data from articles

After the manuscript got reject with permission to resubmit at "Frontiers in Forest and Global Change", Dossa contacted Wyka who agreed to collaborate with his colleague. Thus, they provided a database of 10 000 entries of data comparing trees, lianas, and shrubs (live leaves). Moreover, we have added few more articles data (we found in the literature): Santiago 2010 (senesced leaves), Li et al 2013, Scowcroft 1997 (senesced leaves), Heineman et al 2016, Dossa et al 2016, Jones et al 2019, Jones et al 2020, Dossa et al 2020. Moreover, we go the initial raw data from Roeder et al 2022 (senesced leaves), and finally unpublished data from Dossa. Now, we are able to compute the mean and standard error of the following traits: Nitrogen (N), Phosphorus (P), Calcium (Ca), Magnesium (Mg), Carbon (C), Tannins (condensed and water soluble), and Lignin.

## Input file
```{r}
litterquality<-read_csv(file="Liana_perspective_datab.csv",
                        col_types = list(`Literature Source` = col_character(),
  Kingdom = col_character(),
  Subkingdom = col_character(),
  Division = col_character(),
  Class = col_character(),
  Subclass = col_character(),
  Order = col_character(),
  Family = col_character(),
  Genus = col_character(),
  Species = col_character(),
  `Life Form` = col_character(),
  Latitude = col_double(),
  Longitude = col_double(),
  `MAT(degC)` = col_double(),
  `MAP(mm)` = col_double(),
  Altitude = col_double(),
  Site1 = col_character(),
  Country = col_character(),
  Continent = col_character(),
  ClimRegion = col_character(),
  Tissue = col_character(),
  SLA = col_double(),
  LMA = col_double(),
  Nitrogen = col_double(),
  Phosphorus = col_double(),
  Calcium = col_double(),
  Magnesium = col_double(),
  Carbon = col_double(),
  Tannins_cond = col_double(),
  Tannins_wat = col_double(),
  Lignin = col_double(),
  Cellulose_percent = col_double(),
  Hemicellulose_percent = col_double(),
  Fiber_content_percent = col_double(),
  Tissue_status= col_character()))
```



## Compute chemical traits average per tissue


```{r}
glimpse(litterquality)
litterquality_avg<-litterquality%>%
  rename(Life_form=`Life_form`)%>%
  mutate(Life_form=as.factor(Life_form),
         Tissue=as.factor(Tissue),
         SLA=as.numeric(SLA),
         LMA=as.numeric(LMA),
         Tannins_cond=as.numeric(Tannins_cond),
         Tannins_wat=as.numeric(Tannins_wat),
         Cellulose=Cellulose_percent, 
         Hemi_cellu=Hemicellulose_percent,
         Fiber=Fiber_content_percent,
         Magnesium=as.numeric(Magnesium))%>%
    group_by(Tissue, Life_form)%>%
  dplyr::summarise(mean_SLA=mean(SLA, na.rm=T),
                   SD_SLA=sd(SLA, na.rm=T),
                   mean_LMA=mean(LMA, na.rm=T),
                   SD_LMA=sd(LMA, na.rm=T),
                   mean_C=mean(Carbon, na.rm=T),
                   SD_C=sd(Carbon, na.rm=T),
                   mean_N=mean(Nitrogen, na.rm=T),
                   SD_N=sd(Nitrogen, na.rm=T),
                   mean_P=mean(Phosphorus, na.rm=T),
                   SD_P=sd(Phosphorus, na.rm=T),
                   mean_Mg=mean(Magnesium, na.rm=T),
                   SD_Mg=sd(Magnesium, na.rm=T),
                   mean_Ca=mean(Calcium, na.rm=T),
                   SD_Ca=sd(Calcium, na.rm=T),
                   mean_Lignin=mean(Lignin, na.rm=T),
                   SD_Lignin=sd(Lignin, na.rm=T),
                   mean_Tannins_wat=mean(Tannins_wat, na.rm=T),
                   SD_Tannin_wat=sd(Tannins_wat, na.rm=T),
                   mean_Tannin_cond=mean(Tannins_cond, na.rm=T),
                   SD_Tannin_cond=sd(Tannins_cond, na.rm=T),
                   size=n()
)
litterquality_avg
```


Select Dilleniaceae family and compare their means with the liana mean 

```{r}
glimpse(litterquality)
litterquality_Dil<-litterquality%>%
  filter(str_detect(Family,'Dilleniaceae'))%>%
  group_by(Family, Tissue, Life_form)%>%
  dplyr::summarise(mean_SLA=mean(SLA, na.rm=T),
                   SD_SLA=sd(SLA, na.rm=T),
                   mean_LMA=mean(LMA, na.rm=T),
                   SD_LMA=sd(LMA, na.rm=T),
                   mean_C=mean(Carbon, na.rm=T),
                   SD_C=sd(Carbon, na.rm=T),
                   mean_N=mean(Nitrogen, na.rm=T),
                   SD_N=sd(Nitrogen, na.rm=T),
                   mean_P=mean(Phosphorus, na.rm=T),
                   SD_P=sd(Phosphorus, na.rm=T),
                   mean_Mg=mean(Magnesium, na.rm=T),
                   SD_Mg=sd(Magnesium, na.rm=T),
                   mean_Ca=mean(Calcium, na.rm=T),
                   SD_Ca=sd(Calcium, na.rm=T),
                   mean_Lignin=mean(Lignin, na.rm=T),
                   SD_Lignin=sd(Lignin, na.rm=T),
                   mean_Tannins_wat=mean(Tannins_wat, na.rm=T),
                   SD_Tannin_wat=sd(Tannins_wat, na.rm=T),
                   mean_Tannin_cond=mean(Tannins_cond, na.rm=T),
                   SD_Tannin_cond=sd(Tannins_cond, na.rm=T),
                   size=n()
)
litterquality_Dil
```

A reviewer of the manuscript mentioned about low nutrient in the family of Dilleniaceae. Thus, I subset the family of Dilleniaceae above and compare the nutrients with the average computed for all liana leaves. We found that although our database contains only one species that belongs to this family, its nutrients concentrations are way below the liana means. For instance, Dilleniacea has N = 14.7 vs liana N (23.096 mg g-1), while Dilleniaceae has P = 0.79 vs. liana P (1.63 mg g-1). I could then conclude that it is possible that Dilleniaceae has low nutrient but more data are needed to confirm such trend.


## Compute chemical traits average per tissue with sample size per element
One of our co-authors Wyka ask to be more specific about where data come from and give the sample size per chemical element. 

SLA
```{r}
SLA_avg<-litterquality%>%
  select(Tissue, Life_form, SLA)%>%
  filter(!str_detect(SLA,'NA'))%>%
  group_by(Tissue, Life_form)%>%
  dplyr::summarise(mean_SLA=mean(SLA, na.rm=T),
                   SD_SLA=sd(SLA, na.rm=T),
                   size_SLA=n()
)
SLA_avg
```


LMA
```{r}
LMA_avg<-litterquality%>%
  select(Tissue, Life_form, LMA)%>%
  filter(!str_detect(LMA,'NA'))%>%
  group_by(Tissue, Life_form)%>%
  dplyr::summarise(mean_LMA=mean(LMA, na.rm=T),
                   SD_LMA=sd(LMA, na.rm=T),
                   size_LMA=n()
)
LMA_avg
```
Nitrogen
```{r}
Nitrogen_avg<-litterquality%>%
  select(Tissue, Life_form, Nitrogen)%>%
  filter(!str_detect(Nitrogen,'NA'))%>%
  group_by(Tissue, Life_form)%>%
  dplyr::summarise(mean_Nitrogen=mean(Nitrogen, na.rm=T),
                   SD_Nitrogen=sd(Nitrogen, na.rm=T),
                   size_Nitrogen=n()
)
Nitrogen_avg
```

Phosphorus
```{r}
Phosphorus_avg<-litterquality%>%
  select(Tissue, Life_form, Phosphorus)%>%
  filter(!str_detect(Phosphorus,'NA'))%>%
  group_by(Tissue, Life_form)%>%
  dplyr::summarise(mean_Phosphorus=mean(Phosphorus, na.rm=T),
                   SD_Phosphorus=sd(Phosphorus, na.rm=T),
                   size_Phosphorus=n()
)
Phosphorus_avg
```


Carbon
```{r}
Carbon_avg<-litterquality%>%
  select(Tissue, Life_form, Carbon)%>%
  filter(!str_detect(Carbon,'NA'))%>%
  group_by(Tissue, Life_form)%>%
  dplyr::summarise(mean_Carbon=mean(Carbon, na.rm=T),
                   SD_Carbon=sd(Carbon, na.rm=T),
                   size_Carbon=n()
)
Carbon_avg
```

Lignin

```{r}
Lignin_avg<-litterquality%>%
  select(Tissue, Life_form, Lignin)%>%
  filter(!str_detect(Lignin,'NA'))%>%
  group_by(Tissue, Life_form)%>%
  dplyr::summarise(mean_Lignin=mean(Lignin, na.rm=T),
                   SD_Lignin=sd(Lignin, na.rm=T),
                   size_Lignin=n()
)
Lignin_avg
```

```{r}
Tannins_wat_avg<-litterquality%>%
  select(Tissue, Life_form, Tannins_wat)%>%
  filter(!str_detect(Tannins_wat,'NA'))%>%
  group_by(Tissue, Life_form)%>%
  dplyr::summarise(mean_Tannin_wat=mean(Tannins_wat, na.rm=T),
                   SD_Tannin_wat=sd(Tannins_wat, na.rm=T),
                   size_Tannin_wat=n()
)
Tannins_wat_avg
```
```{r}
Tannins_cond_avg<-litterquality%>%
  select(Tissue, Life_form, Tannins_cond)%>%
  filter(!str_detect(Tannins_cond,'NA'))%>%
  group_by(Tissue, Life_form)%>%
  dplyr::summarise(mean_Tannin_cond=mean(Tannins_cond, na.rm=T),
                   SD_Tannin_cond=sd(Tannins_cond, na.rm=T),
                   size_Tannin_cond=n()
)
Tannins_cond_avg
```


Magnesium
```{r}
Magnesium_avg<-litterquality%>%
  select(Tissue, Life_form, Magnesium)%>%
  filter(!str_detect(Magnesium,'NA'))%>%
  group_by(Tissue, Life_form)%>%
  dplyr::summarise(mean_Magnesium=mean(Magnesium, na.rm=T),
                   SD_Magnesium=sd(Magnesium, na.rm=T),
                   size_Magnesium=n()
)
Magnesium_avg
```

Calcium
```{r}
Calcium_avg<-litterquality%>%
  select(Tissue, Life_form, Calcium)%>%
  filter(!str_detect(Calcium,'NA'))%>%
  group_by(Tissue, Life_form)%>%
  dplyr::summarise(mean_Calcium=mean(Calcium, na.rm=T),
                   SD_Calcium=sd(Calcium, na.rm=T),
                   size_Calcium=n()
)
Calcium_avg
```



```{r}
litterquality_avg2<-litterquality%>%
  group_by(Tissue, Life_form)%>%
  dplyr::summarise(mean_SLA=mean(SLA, na.rm=T),
                   SD_SLA=sd(SLA, na.rm=T),
                   size_SLA=n(),
                   mean_LMA=mean(LMA, na.rm=T),
                   SD_LMA=sd(LMA, na.rm=T),
                   size_LMA=n(),
                   mean_C=mean(Carbon, na.rm=T),
                   SD_C=sd(Carbon, na.rm=T),
                   size_C=n(),
                   mean_N=mean(Nitrogen, na.rm=T),
                   SD_N=sd(Nitrogen, na.rm=T),
                   size_N=n(),
                   mean_P=mean(Phosphorus, na.rm=T),
                   SD_P=sd(Phosphorus, na.rm=T),
                   size_P=n(),
                   mean_Mg=mean(Magnesium, na.rm=T),
                   SD_Mg=sd(Magnesium, na.rm=T),
                   size_Mg=n(),
                   mean_Ca=mean(Calcium, na.rm=T),
                   SD_Ca=sd(Calcium, na.rm=T),
                   size_Ca=n(),
                   mean_Lignin=mean(Lignin, na.rm=T),
                   SD_Lignin=sd(Lignin, na.rm=T),
                   size_Lignin=n(),
                   mean_Tannins_wat=mean(Tannins_wat, na.rm=T),
                   SD_Tannin_wat=sd(Tannins_wat, na.rm=T),
                   size_Tannin_wat=n(),
                   mean_Tannin_cond=mean(Tannins_cond, na.rm=T),
                   SD_Tannin_cond=sd(Tannins_cond, na.rm=T),
                   size_Tannin_cond=n()
)
litterquality_avg2
```


## Carry out t.test on different chemical and per organ

This was asked by Richard Corlett and he suggests to carry out these analyses and report the limitations of the data anyway.

SLA
Leaf
```{r}
SLA<-litterquality%>%
  select(Tissue, Life_form, SLA)%>%
  filter(!str_detect(SLA,'NA'))
SLA
t.test(data=SLA, SLA~Life_form)
```


LMA
Leaf
```{r}
LMA<-litterquality%>%
  select(Tissue, Life_form, LMA)%>%
  filter(!str_detect(LMA,'NA'))
LMA
t.test(data=LMA, LMA~Life_form)
```
Nitrogen
Leaf
```{r}
Nitrogen_leaf<-litterquality%>%
  select(Tissue, Life_form, Nitrogen)%>%
  filter(!str_detect(Nitrogen,'NA'))%>%
  filter(str_detect(Tissue,'leaf'))
Nitrogen_leaf
t.test(data=Nitrogen_leaf, Nitrogen~Life_form)
```

bark

```{r}
Nitrogen_bark<-litterquality%>%
  select(Tissue, Life_form, Nitrogen)%>%
  filter(!str_detect(Nitrogen,'NA'))%>%
  filter(str_detect(Tissue,'bark'))
Nitrogen_bark
t.test(data=Nitrogen_bark, Nitrogen~Life_form)
```

wood

```{r}
Nitrogen_wood<-litterquality%>%
  select(Tissue, Life_form, Nitrogen)%>%
  filter(!str_detect(Nitrogen,'NA'))%>%
  filter(str_detect(Tissue,'wood'))
Nitrogen_wood
t.test(data=Nitrogen_wood, Nitrogen~Life_form)
```
Phosphorus

leaf
```{r}
Phosphorus_leaf<-litterquality%>%
  select(Tissue, Life_form, Phosphorus)%>%
  filter(!str_detect(Phosphorus,'NA'))%>%
  filter(str_detect(Tissue,'leaf'))
Phosphorus_leaf
t.test(data=Phosphorus_leaf, Phosphorus~Life_form)
```

bark
```{r}
Phosphorus_bark<-litterquality%>%
  select(Tissue, Life_form, Phosphorus)%>%
  filter(!str_detect(Phosphorus,'NA'))%>%
  filter(str_detect(Tissue,'bark'))
Phosphorus_bark
t.test(data=Phosphorus_bark, Phosphorus~Life_form)
```

wood
```{r}
Phosphorus_wood<-litterquality%>%
  select(Tissue, Life_form, Phosphorus)%>%
  filter(!str_detect(Phosphorus,'NA'))%>%
  filter(str_detect(Tissue,'wood'))
Phosphorus_wood
t.test(data=Phosphorus_wood, Phosphorus~Life_form)
```
Carbon

leaf
```{r}
Carbon_leaf<-litterquality%>%
  select(Tissue, Life_form, Carbon)%>%
  filter(!str_detect(Carbon,'NA'))%>%
  filter(str_detect(Tissue,'leaf'))
Carbon_leaf
t.test(data=Carbon_leaf, Carbon~Life_form)
```

bark
```{r}
Carbon_bark<-litterquality%>%
  select(Tissue, Life_form, Carbon)%>%
  filter(!str_detect(Carbon,'NA'))%>%
  filter(str_detect(Tissue,'bark'))
Carbon_bark
t.test(data=Carbon_bark, Carbon~Life_form)
```
wood
```{r}
Carbon_wood<-litterquality%>%
  select(Tissue, Life_form, Carbon)%>%
  filter(!str_detect(Carbon,'NA'))%>%
  filter(str_detect(Tissue,'wood'))
Carbon_wood
t.test(data=Carbon_wood, Carbon~Life_form)
```

Lignin

leaf
```{r}
Lignin_leaf<-litterquality%>%
  select(Tissue, Life_form, Lignin)%>%
  filter(!str_detect(Lignin,'NA'))%>%
  filter(str_detect(Tissue,'leaf'))
Lignin_leaf
t.test(data=Lignin_leaf, Lignin~Life_form)
```

bark
```{r}
Lignin_bark<-litterquality%>%
  select(Tissue, Life_form, Lignin)%>%
  filter(!str_detect(Lignin,'NA'))%>%
  filter(str_detect(Tissue,'bark'))
Lignin_bark
t.test(data=Lignin_bark, Lignin~Life_form)
```

wood
```{r}
Lignin_wood<-litterquality%>%
  select(Tissue, Life_form, Lignin)%>%
  filter(!str_detect(Lignin,'NA'))%>%
  filter(str_detect(Tissue,'wood'))
Lignin_wood
t.test(data=Lignin_wood, Lignin~Life_form)
```

Soluble tannins

leaf
```{r}
glimpse(litterquality)
Tannins_wat_leaf<-litterquality%>%
  select(Tissue, Life_form, Tannins_wat)%>%
  filter(!str_detect(Tannins_wat,'NA'))%>%
  filter(str_detect(Tissue,'leaf'))
Tannins_wat_leaf
t.test(data=Tannins_wat_leaf, Tannins_wat~Life_form)
```

Condensed tannins
bark

```{r}
Tannins_cond_bark<-litterquality%>%
  select(Tissue, Life_form, Tannins_cond)%>%
  filter(!str_detect(Tannins_cond,'NA'))%>%
  filter(str_detect(Tissue,'bark'))
Tannins_cond_bark
t.test(data=Tannins_cond_bark, Tannins_cond~Life_form)
```

wood

```{r}
Tannins_cond_wood<-litterquality%>%
  select(Tissue, Life_form, Tannins_cond)%>%
  filter(!str_detect(Tannins_cond,'NA'))%>%
  filter(str_detect(Tissue,'wood'))
Tannins_cond_wood
t.test(data=Tannins_cond_wood, Tannins_cond~Life_form)
```



Calcium

leaf
```{r}
Calcium_leaf<-litterquality%>%
  select(Tissue, Life_form, Calcium)%>%
  filter(!str_detect(Calcium,'NA'))%>%
  filter(str_detect(Tissue,'leaf'))
Calcium_leaf
t.test(data=Calcium_leaf, Calcium~Life_form)
```

bark
```{r}
Calcium_bark<-litterquality%>%
  select(Tissue, Life_form, Calcium)%>%
  filter(!str_detect(Calcium,'NA'))%>%
  filter(str_detect(Tissue,'bark'))
Calcium_bark
t.test(data=Calcium_bark, Calcium~Life_form)
```

wood
```{r}
Calcium_wood<-litterquality%>%
  select(Tissue, Life_form, Calcium)%>%
  filter(!str_detect(Calcium,'NA'))%>%
  filter(str_detect(Tissue,'wood'))
Calcium_wood
t.test(data=Calcium_wood, Calcium~Life_form)
```

Magnesium

leaf
```{r}
Magnesium_leaf<-litterquality%>%
  select(Tissue, Life_form, Magnesium)%>%
  filter(!str_detect(Magnesium,'NA'))%>%
  filter(str_detect(Tissue,'leaf'))
Magnesium_leaf
t.test(data=Magnesium_leaf, Magnesium~Life_form)
```

bark
```{r}
Magnesium_bark<-litterquality%>%
  select(Tissue, Life_form, Magnesium)%>%
  filter(!str_detect(Magnesium,'NA'))%>%
  filter(str_detect(Tissue,'bark'))
Magnesium_bark
t.test(data=Magnesium_bark, Magnesium~Life_form)
```

wood
```{r}
Magnesium_wood<-litterquality%>%
  select(Tissue, Life_form, Magnesium)%>%
  filter(!str_detect(Magnesium,'NA'))%>%
  filter(str_detect(Tissue,'wood'))
Magnesium_wood
t.test(data=Magnesium_wood, Magnesium~Life_form)
```

One co-author asked to separate live leaves chemistry from senesced ones.

## Input file
```{r}
litterquality<-read_csv(file="Liana_perspective_datab.csv",
                        col_types = list(`Literature Source` = col_character(),
  Kingdom = col_character(),
  Subkingdom = col_character(),
  Division = col_character(),
  Class = col_character(),
  Subclass = col_character(),
  Order = col_character(),
  Family = col_character(),
  Genus = col_character(),
  Species = col_character(),
  `Life Form` = col_character(),
  Latitude = col_double(),
  Longitude = col_double(),
  `MAT(degC)` = col_double(),
  `MAP(mm)` = col_double(),
  Altitude = col_double(),
  Site1 = col_character(),
  Country = col_character(),
  Continent = col_character(),
  ClimRegion = col_character(),
  Tissue = col_character(),
  SLA = col_double(),
  LMA = col_double(),
  Nitrogen = col_double(),
  Phosphorus = col_double(),
  Calcium = col_double(),
  Magnesium = col_double(),
  Carbon = col_double(),
  Tannins_cond = col_double(),
  Tannins_wat = col_double(),
  Lignin = col_double(),
  Cellulose_percent = col_double(),
  Hemicellulose_percent = col_double(),
  Fiber_content_percent = col_double(),
  Tissue_status= col_character()))
```



## Compute chemical traits average per tissue


```{r}
glimpse(litterquality)
litterquality_avg<-litterquality%>%
  rename(Life_form=`Life_form`)%>%
  mutate(Life_form=as.factor(Life_form),
         Tissue=as.factor(Tissue),
         SLA=as.numeric(SLA),
         LMA=as.numeric(LMA),
         Tannins_cond=as.numeric(Tannins_cond),
         Tannins_wat=as.numeric(Tannins_wat),
         Cellulose=Cellulose_percent, 
         Hemi_cellu=Hemicellulose_percent,
         Fiber=Fiber_content_percent,
         Magnesium=as.numeric(Magnesium))%>%
    group_by(Tissue, Life_form, Tissue_status)%>%
  dplyr::summarise(mean_SLA=mean(SLA, na.rm=T),
                   SD_SLA=sd(SLA, na.rm=T),
                   mean_LMA=mean(LMA, na.rm=T),
                   SD_LMA=sd(LMA, na.rm=T),
                   mean_C=mean(Carbon, na.rm=T),
                   SD_C=sd(Carbon, na.rm=T),
                   mean_N=mean(Nitrogen, na.rm=T),
                   SD_N=sd(Nitrogen, na.rm=T),
                   mean_P=mean(Phosphorus, na.rm=T),
                   SD_P=sd(Phosphorus, na.rm=T),
                   mean_Mg=mean(Magnesium, na.rm=T),
                   SD_Mg=sd(Magnesium, na.rm=T),
                   mean_Ca=mean(Calcium, na.rm=T),
                   SD_Ca=sd(Calcium, na.rm=T),
                   mean_Lignin=mean(Lignin, na.rm=T),
                   SD_Lignin=sd(Lignin, na.rm=T),
                   mean_Tannins_wat=mean(Tannins_wat, na.rm=T),
                   SD_Tannin_wat=sd(Tannins_wat, na.rm=T),
                   mean_Tannin_cond=mean(Tannins_cond, na.rm=T),
                   SD_Tannin_cond=sd(Tannins_cond, na.rm=T),
                   size=n()
)
litterquality_avg
```


Average with tissue status (live or senesced)

```{r}
litterquality_avg3<-litterquality%>%
  group_by(Tissue, Life_form, Tissue_status)%>%
  dplyr::summarise(mean_SLA=mean(SLA, na.rm=T),
                   SD_SLA=sd(SLA, na.rm=T),
                   size_SLA=n(),
                   mean_LMA=mean(LMA, na.rm=T),
                   SD_LMA=sd(LMA, na.rm=T),
                   size_LMA=n(),
                   mean_C=mean(Carbon, na.rm=T),
                   SD_C=sd(Carbon, na.rm=T),
                   size_C=n(),
                   mean_N=mean(Nitrogen, na.rm=T),
                   SD_N=sd(Nitrogen, na.rm=T),
                   size_N=n(),
                   mean_P=mean(Phosphorus, na.rm=T),
                   SD_P=sd(Phosphorus, na.rm=T),
                   size_P=n(),
                   mean_Mg=mean(Magnesium, na.rm=T),
                   SD_Mg=sd(Magnesium, na.rm=T),
                   size_Mg=n(),
                   mean_Ca=mean(Calcium, na.rm=T),
                   SD_Ca=sd(Calcium, na.rm=T),
                   size_Ca=n(),
                   mean_Lignin=mean(Lignin, na.rm=T),
                   SD_Lignin=sd(Lignin, na.rm=T),
                   size_Lignin=n(),
                   mean_Tannins_wat=mean(Tannins_wat, na.rm=T),
                   SD_Tannin_wat=sd(Tannins_wat, na.rm=T),
                   size_Tannin_wat=n(),
                   mean_Tannin_cond=mean(Tannins_cond, na.rm=T),
                   SD_Tannin_cond=sd(Tannins_cond, na.rm=T),
                   size_Tannin_cond=n()
)
litterquality_avg3
```


## Carry out t.test on different chemical for leaf do for live and senesced ones

### Live leaves
SLA
Leaf live
```{r}
SLA_live<-litterquality%>%
  select(Tissue, Life_form, Tissue_status,SLA)%>%
  filter(!str_detect(SLA,'NA'))%>%
  filter(str_detect(Tissue,'leaf'))%>%
  filter(str_detect(Tissue_status, "live"))
SLA_live
t.test(data=SLA_live, SLA~Life_form)
# SLA higher for live liana leaves than live tree leaves (t=7.781, df=178.89, p<0.001)
```


LMA
Leaf
```{r}
LMA_live<-litterquality%>%
  select(Tissue, Life_form,Tissue_status, LMA)%>%
  filter(!str_detect(LMA,'NA'))%>%
  filter(str_detect(Tissue,'leaf'))%>%
  filter(str_detect(Tissue_status, "live"))
LMA_live
t.test(data=LMA_live, LMA~Life_form)
# Lower LMA in live liana leaves than live tree leaves (t=-9.271, df=308.54, p<0.001)
```

Nitrogen
Leaf live
```{r}
Nitrogen_leaf_live<-litterquality%>%
  select(Tissue, Life_form,Tissue_status, Nitrogen)%>%
  filter(!str_detect(Nitrogen,'NA'))%>%
  filter(str_detect(Tissue,'leaf'))%>%
  filter(str_detect(Tissue_status, "live"))
Nitrogen_leaf_live
t.test(data=Nitrogen_leaf_live, Nitrogen~Life_form)

# Nitrogen higher in live liana leaves than live tree leaves (t= 6.579, df=200.77, p< 0.001)
```


Phosphorus

leaf live
```{r}
Phosphorus_leaf_live<-litterquality%>%
  select(Tissue, Life_form,Tissue_status, Phosphorus)%>%
  filter(!str_detect(Phosphorus,'NA'))%>%
  filter(str_detect(Tissue,'leaf'))%>%
  filter(str_detect(Tissue_status,'live'))
Phosphorus_leaf_live
t.test(data=Phosphorus_leaf_live, Phosphorus~Life_form)

# Phosphorus higher in live liana leaves than live tree leaves (t= 3.5446, df=136.69, p< 0.001)
```

Carbon

leaf live
```{r}
Carbon_leaf_live<-litterquality%>%
  select(Tissue, Life_form, Tissue_status, Carbon)%>%
  filter(!str_detect(Carbon,'NA'))%>%
  filter(str_detect(Tissue,'leaf'))%>%
  filter(str_detect(Tissue_status,'live'))
Carbon_leaf_live

# No information available so no test.
```

Lignin

leaf live
```{r}
Lignin_leaf_live<-litterquality%>%
  select(Tissue, Life_form,Tissue_status, Lignin)%>%
  filter(!str_detect(Lignin,'NA'))%>%
  filter(str_detect(Tissue,'leaf'))%>%
  filter(str_detect(Tissue_status,'live'))
Lignin_leaf_live
# No information available so no test.

```

Soluble tannins

leaf live
```{r}
glimpse(litterquality)
Tannins_wat_leaf_live<-litterquality%>%
  select(Tissue, Life_form, Tissue_status, Tannins_wat)%>%
  filter(!str_detect(Tannins_wat,'NA'))%>%
  filter(str_detect(Tissue,'leaf'))%>%
  filter(str_detect(Tissue_status, "live"))
Tannins_wat_leaf_live
# No information available so no test.

```
Calcium

leaf
```{r}
Calcium_leaf_live<-litterquality%>%
  select(Tissue, Life_form, Tissue_status, Calcium)%>%
  filter(!str_detect(Calcium,'NA'))%>%
  filter(str_detect(Tissue,'leaf'))%>%
  filter(str_detect(Tissue_status, "live"))
Calcium_leaf_live
t.test(data=Calcium_leaf_live, Calcium~Life_form)

# Calcium higher in live liana leaves than live tree leaves (t= 5.0489, df=91.74, p< 0.001)
```


Magnesium

leaf live
```{r}
Magnesium_leaf_live<-litterquality%>%
  select(Tissue, Life_form, Tissue_status, Magnesium)%>%
  filter(!str_detect(Magnesium,'NA'))%>%
  filter(str_detect(Tissue,'leaf'))%>%
  filter(str_detect(Tissue_status, "live"))
Magnesium_leaf_live
# No information available so no test.
```


### senesced leaves
SLA
Leaf senesced
```{r}
SLA_senesced<-litterquality%>%
  select(Tissue, Life_form, Tissue_status,SLA)%>%
  filter(!str_detect(SLA,'NA'))%>%
  filter(str_detect(Tissue,'leaf'))%>%
  filter(str_detect(Tissue_status, "senesced"))
SLA_senesced
t.test(data=SLA_senesced, SLA~Life_form)
# SLA higher for senesced liana leaves than senesced tree leaves (t=4.6501, df=41.941, p<0.001)
```


LMA
Leaf
```{r}
LMA_senesced<-litterquality%>%
  select(Tissue, Life_form,Tissue_status, LMA)%>%
  filter(!str_detect(LMA,'NA'))%>%
  filter(str_detect(Tissue,'leaf'))%>%
  filter(str_detect(Tissue_status, "senesced"))
LMA_senesced
# No data available on this
```

Nitrogen
Leaf senesced
```{r}
Nitrogen_leaf_senesced<-litterquality%>%
  select(Tissue, Life_form,Tissue_status, Nitrogen)%>%
  filter(!str_detect(Nitrogen,'NA'))%>%
  filter(str_detect(Tissue,'leaf'))%>%
  filter(str_detect(Tissue_status, "senesced"))
Nitrogen_leaf_senesced
t.test(data=Nitrogen_leaf_senesced, Nitrogen~Life_form)

# Nitrogen higher in senesced liana leaves than senesced tree leaves (t= 2.4569, df=34.595, p< 0.0191)
```


Phosphorus

leaf senesced
```{r}
Phosphorus_leaf_senesced<-litterquality%>%
  select(Tissue, Life_form,Tissue_status, Phosphorus)%>%
  filter(!str_detect(Phosphorus,'NA'))%>%
  filter(str_detect(Tissue,'leaf'))%>%
  filter(str_detect(Tissue_status,'senesced'))
Phosphorus_leaf_senesced
t.test(data=Phosphorus_leaf_senesced, Phosphorus~Life_form)

# No difference between Phosphorus in senesced liana leaves than senesced tree leaves (t= 0.0317, df=40.829, p< 0.9748)
```

Carbon

leaf senesced
```{r}
Carbon_leaf_senesced<-litterquality%>%
  select(Tissue, Life_form, Tissue_status, Carbon)%>%
  filter(!str_detect(Carbon,'NA'))%>%
  filter(str_detect(Tissue,'leaf'))%>%
  filter(str_detect(Tissue_status,'senesced'))
Carbon_leaf_senesced
t.test(data=Carbon_leaf_senesced, Carbon~Life_form)

# No difference in carbon of senesced liana leaves and senesced tree leaves.
```

Lignin

leaf senesced
```{r}
Lignin_leaf_senesced<-litterquality%>%
  select(Tissue, Life_form,Tissue_status, Lignin)%>%
  filter(!str_detect(Lignin,'NA'))%>%
  filter(str_detect(Tissue,'leaf'))%>%
  filter(str_detect(Tissue_status,'senesced'))
Lignin_leaf_senesced
t.test(data=Lignin_leaf_senesced, Lignin~Life_form)
# No difference between senesced liana leaves and senesced leaves (t=0.3790, df=43.96, p=0.7065).

```

Soluble tannins

leaf senesced
```{r}
glimpse(litterquality)
Tannins_wat_leaf_senesced<-litterquality%>%
  select(Tissue, Life_form, Tissue_status, Tannins_wat)%>%
  filter(!str_detect(Tannins_wat,'NA'))%>%
  filter(str_detect(Tissue,'leaf'))%>%
  filter(str_detect(Tissue_status, "senesced"))
Tannins_wat_leaf_senesced
t.test(data=Tannins_wat_leaf_senesced, Tannins_wat~Life_form)
# No difference in Tannin water soluble between senesced liana leaves and three scenesced tree leaves (t=-0.4708, df=37.302, p=0.6405). 

```
Calcium

leaf
```{r}
Calcium_leaf_senesced<-litterquality%>%
  select(Tissue, Life_form, Tissue_status, Calcium)%>%
  filter(!str_detect(Calcium,'NA'))%>%
  filter(str_detect(Tissue,'leaf'))%>%
  filter(str_detect(Tissue_status, "senesced"))
Calcium_leaf_senesced
t.test(data=Calcium_leaf_senesced, Calcium~Life_form)

# No difference in Calcium between senesced liana leaves and senesced tree leaves (t= 1.3675, df=40.647, p=0.179). 
```


Magnesium

leaf senesced
```{r}
Magnesium_leaf_senesced<-litterquality%>%
  select(Tissue, Life_form, Tissue_status, Magnesium)%>%
  filter(!str_detect(Magnesium,'NA'))%>%
  filter(str_detect(Tissue,'leaf'))%>%
  filter(str_detect(Tissue_status, "senesced"))
Magnesium_leaf_senesced
t.test(data=Magnesium_leaf_senesced, Magnesium~Life_form)

# No difference in Magnesium between senesced liana leaves and tree ones (t=1.3074, df=37.315, p=0.1991).
```